Supplemental Information: Aligning marine species range data to better serve science and conservation

### 2016-08-15 case_study_ohi/ohi_paired_maps.Rmd

``` {r create_box_plots}
dir_fig <- '~/github/IUCN-AquaMaps/figures'
source(file.path(dir_git, 'data_explore/data_explore_fxns.R'))

@@

print(bplot_abs)
ggsave(file.path(dir_fig, 'fig3a_boxplot_paired_maps.png'),
       height = 4.5, width = 8.7, units = 'cm', dpi = 600)
ggsave(file.path(dir_fig, 'fig3a_boxplot_paired_maps.tif'), device = 'tiff',
       height = 4.5, width = 8.7, units = 'cm', dpi = 600)

>>>>

``` {r create_violin_plots}
dir_fig <- '~/github/IUCN-AquaMaps/figures'
source(file.path(dir_git, 'data_explore/data_explore_fxns.R'))

@@

print(bplot_abs)
ggsave(file.path(dir_fig, 'fig3a_violinplot_paired_maps.png'),
       height = 4.5, width = 8.7, units = 'cm', dpi = 600)
ggsave(file.path(dir_fig, 'fig3a_violinplot_paired_maps.tif'), device = 'tiff',
       height = 4.5, width = 8.7, units = 'cm', dpi = 600)

### 2016-08-15 ms_0_abst_sig.Rmd

Species distribution data provide the foundation for a wide range of ecological research studies and conservation management decisions, yet most species ranges remain unknown, and existing range maps often suffer from data limitations and inconsistencies.  AquaMaps and the International Union for Conservation of Nature (IUCN) are two distinct efforts to map marine species distributions at a global scale.  Together these databases represent `r format(n_spp_all, big.mark = ',')` species (`r format(p_spp_am, big.mark = ',')`% within AquaMaps, `r format(p_spp_iucn, big.mark = ',')`% within IUCN), with only `r format(n_spp_both, big.mark = ',')` shared species.  Here we examine differences in predicted species ranges between the two datasets and find that these misalignments mainly result from divergent methodologies that introduce differing frequencies of commission and omission errors.  We illustrate the scientific and management implications of these differences by repeating two recent applications - an assessment of global biodiversity within the Ocean Health Index and a global analysis of gaps in coverage of marine protected areas - and find significantly different results depending on how the two datasets were used.  Until a single, highly accurate dataset of global marine species ranges becomes available, understanding the implications of dataset differences for conservation planning and decision-making remains essential.

>>>>

Species distribution data provide the foundation for a wide range of ecological research studies and conservation management decisions, yet most species ranges remain unknown, and existing range maps often suffer from data limitations and inconsistencies.  AquaMaps and the International Union for Conservation of Nature (IUCN) are two distinct efforts to map marine species distributions at a global scale.  Together these databases represent `r format(n_spp_all, big.mark = ',')` species (`r format(p_spp_am, big.mark = ',')`% within AquaMaps, `r format(p_spp_iucn, big.mark = ',')`% within IUCN), with only `r format(n_spp_both, big.mark = ',')` shared species.  Here we examine differences in predicted species ranges between the two datasets and find that these misalignments mainly result from divergent methodologies that introduce differing frequencies of commission and omission errors.  We illustrate the scientific and management implications of these differences by repeating two recent applications - an assessment of global biodiversity within the Ocean Health Index and a global analysis of gaps in coverage of marine protected areas - and find significantly different results depending on how the two datasets are used.  Until a single, highly accurate dataset of global marine species ranges becomes available, understanding the implications of dataset differences for conservation planning and decision-making remains essential.

### 2016-08-15 ms_1_intro.Rmd

<!--
The two datasets ostensibly describe the same information, but significant differences in methodology and intent could lead to dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations. 
Discuss "truth" and uncertainty in species distributions.
There is an objective "truth"; 
-->
Neither dataset claims to represent the "truth" of a species' spatial distribution, but rather each presents one understanding of the "truth" - one database relying primarily on expert opinion and the other on model predictions.  Uncertainties inherent in the method and intent inevitably drive differences in the range predictions made by each dataset; significant differences could indicate dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations.

Inaccurate indications of presence or absence could lead to ineffective marine reserve systems and management plans [@rondinini_tradeoffs_2006; @jetz_ecological_2008].  Importantly, biases in taxonomic or spatial coverage within a dataset could shift management and conservation actions away from places or species that are most in need.

>>>>

___sentence about SAUP - how can we rephrase the "two global-scale repositories" sentence to exclude it? "broad taxonomic coverage"?  something about not focusing just on economically important species? (is that a valid assumption about SAUP)___

___or: "Several global-scale repositories predict ... In this paper we focus on two datasets that cover a broad taxonomic range: ..."___

<!--
The two datasets ostensibly describe the same information, but significant differences in methodology and intent could lead to dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations. 
Discuss "truth" and uncertainty in species distributions.
There is an objective "truth"; 
-->

Neither dataset claims to represent the "truth" of a species' spatial distribution, but rather each presents one understanding of the "truth" - one database relying primarily on expert opinion and the other on model predictions.  Uncertainties inherent in the method and intent inevitably drive differences in the range predictions made by each dataset; significant differences could indicate dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations.  Inaccurate indications of presence or absence could lead to ineffective marine reserve systems and management plans [@rondinini_tradeoffs_2006; @jetz_ecological_2008].  Importantly, biases in taxonomic or spatial coverage within a dataset could shift management and conservation actions away from places or species that are most in need.

### 2016-08-15 ms_2_methods.Rmd

The IUCN publishes species range maps based on expert input of spatial boundaries of a given species' "limits of distribution" [@iucn_iucn_2015] - essentially a refined extent of occurrence (EOO), based on observation records and informed by expert understanding of species' range and habitat preferences.  In contrast, AquaMaps models species distribution based on environmental preferences (e.g., temperature, depth, salinity) deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a 0.5 degree grid, creating a global map of "probability of occurrence" for each species [@kaschner_mapping_2006; @ready_predicting_2010].  Studies using AquaMaps data frequently define "presence" by applying a probability threshold, e.g., previously, the Ocean Health Index defined presence as 40% or greater probability of occurrence [@halpern_index_2012; @halpern_patterns_2015].  In the 2016 OHI assessment, the presence threshold was eliminated, to better approximate the "extent of occurrence" generally indicated by IUCN maps, since the two datasets are used together. [@REF???]

>>>>

The IUCN publishes species range maps based on expert input of spatial boundaries of a given species' "limits of distribution" [@iucn_iucn_2015] - essentially a refined extent of occurrence (EOO), based on observation records and informed by expert understanding of species' range and habitat preferences.  In contrast, AquaMaps models species distribution based on environmental preferences (e.g., temperature, depth, salinity) deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a 0.5 degree grid, creating a global map of suitable habitat.  The model masks habitat to certain FAO Major Fishing Areas as a rough cut of geographic range, resulting in a "probability of occurrence" for each species.  Of the resulting maps, 1296 (5.7%) have been further refined under expert review [@kaschner_mapping_2006; @ready_predicting_2010].  Studies using AquaMaps data frequently define "presence" by applying a probability threshold, e.g., previously, the Ocean Health Index defined presence as 40% or greater probability of occurrence [@halpern_index_2012; @halpern_patterns_2015].  In the 2016 OHI assessment, the presence threshold was eliminated, to better approximate the "extent of occurrence" generally indicated by IUCN maps, since the two datasets are used together.

___short paragraph about other datasets we didn't include, i.e. SAUP?___

### 2016-08-15 ms_6_supplemental.Rmd

# Supplemental Information

@@

Text: Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference).
Figures: Provide a brief legend for each supporting figure after the supporting text. Provide figure images in TIFF, EPS, high-resolution PDF, JPEG, or GIF format; figures may not be embedded in manuscript text. When saving TIFF files, use only LZW compression; do not use JPEG compression. Do not save figure numbers, legends, or author names as part of the image. Composite figures must be preassembled.
3D figures: Supply a composable U3D or PRC file so that it may be edited and composed. Authors may submit a PDF file but please note it will be published in raw format and will not be edited or composed.
Tables: Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference); include only one table per file. Do not use tabs or spaces to separate columns in Word tables.
Datasets: Supply Excel (.xls), RTF, or PDF files. This file type will be published in raw format and will not be edited or composed.
Movies: Supply Audio Video Interleave (avi), Quicktime (mov), Windows Media (wmv), animated GIF (gif), or MPEG files and submit a brief legend for each movie in a Word or RTF file. All movies should be submitted at the desired reproduction size and length. Movies should be no more than 10 MB in size.
Still images: Authors must provide a still image from each video file. Supply TIFF, EPS, high-resolution PDF, JPEG, or GIF files.
Appendices: PNAS prefers that authors submit individual source files to ensure readability. If this is not possible, supply a single PDF file that contains all of the SI associated with the paper. This file type will be published in raw format and will not be edited or composed.

-->

<!--## Data processing - should this go in methods???

@@

### AquaMaps

AquaMaps data was provided by AquaMaps directly via ftp. Three SQL files were sent that contained (1) list of all `r nrow(spp_am)` species and their taxonomic information, (2) information for all (259,200) half degree cells used for range mapping and (3) the probability of occurrence per cell for all `r nrow(spp_am)` species. These were extracted into .csv form using R Statistical Software.

### IUCN

As of December 2015, IUCN had published species distribution maps for `r nrow(spp_iucn_maps)` marine species across 24 taxonomic groups.  For this analysis, we did not consider IUCN range maps for bird species, as those data are hosted separately by BirdLife International.

Each shapefile contains polygon limits of distribution indicating regions of presence/absence, additional attributes provide information on extant/extinct ranges, native/introduced ranges, and seasonality.-->

## S1: sample maps from each quadrant

@@

>  __Fig. S2.__ Breaking down the quadrants by IUCN extinction risk categories (Fig. S2), we found little support for our hypothesis that maps for species with higher extinction risk tend to be better aligned between the two datasets, perhaps correlated to increased expert scrutiny.  Does higher perceived risk lead to increased attention, and thus better understanding of species distribution? Or conversely, does increased attention to species distribution reveal more species at risk?  Likely both mechanisms are at play on a case-by-case basis, depending on the species' taxon and region. _does this argument bear up to closer scrutiny? CR isn't dominated by Q1 any more_

### S3: Coral depth map

@@

<!--
Oculina varicosa - Ivory tree coral ![oculina_pic.jpg](figures/oculina/oculina_pic.jpg)
-->

>  __Fig. S3.__ IUCN-defined range of _Oculina varicosa_, a typical photosynthetic coral, plotted against a 200 m bathymetry contour. According to IUCN, _O. varicosa_ can be found to depths of 152 m. In waters shallower than 30 m, _O. varicosa_ is zooxanthellate, while an azooxanthellate form can be found in deeper waters.

### S4: AquaMaps threshold examination

>>>>

---
title: "Supplemental Information: Aligning marine species range data to better serve science and conservation"
author: "Casey O'Hara, Jamie C. Afflerbach, Courtney Scarborough, Benjamin S. Halpern"
output:
  word_document: default
  pdf_document:
    number_sections: no
    toc: no
  html_document:
    highlight: haddock
    number_sections: no
    theme: spacelab
    toc: no
---

@@

Text: Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference).
Figures: Provide a brief legend for each supporting figure after the supporting text. Provide figure images in TIFF, EPS, high-resolution PDF, JPEG, or GIF format; figures may not be embedded in manuscript text. When saving TIFF files, use only LZW compression; do not use JPEG compression. Do not save figure numbers, legends, or author names as part of the image. Composite figures must be preassembled.
3D figures: Supply a composable U3D or PRC file so that it may be edited and composed. Authors may submit a PDF file but please note it will be published in raw format and will not be edited or composed.
Tables: Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference); include only one table per file. Do not use tabs or spaces to separate columns in Word tables.
Datasets: Supply Excel (.xls), RTF, or PDF files. This file type will be published in raw format and will not be edited or composed.
Movies: Supply Audio Video Interleave (avi), Quicktime (mov), Windows Media (wmv), animated GIF (gif), or MPEG files and submit a brief legend for each movie in a Word or RTF file. All movies should be submitted at the desired reproduction size and length. Movies should be no more than 10 MB in size.
Still images: Authors must provide a still image from each video file. Supply TIFF, EPS, high-resolution PDF, JPEG, or GIF files.
Appendices: PNAS prefers that authors submit individual source files to ensure readability. If this is not possible, supply a single PDF file that contains all of the SI associated with the paper. This file type will be published in raw format and will not be edited or composed.
-->
<!--
# Data sources and methods

@@

## AquaMaps

AquaMaps data was provided by AquaMaps directly via ftp. Three SQL files were sent that contained (1) list of all 22,889 modeled species and their taxonomic information, (2) information for all (259,200) half degree cells used for range mapping and (3) the probability of occurrence per cell for all 22,889 species. These were extracted into .csv form using R Statistical Software.

## IUCN

As of December 2015, IUCN had published species distribution maps for XXX marine species across 24 taxonomic groups.  For this analysis, we did not consider IUCN range maps for bird species, as those data are hosted separately by BirdLife International.

Each shapefile contains spatial polygons outlining limits of distribution for each species, indicating regions of presence/absence; additional attributes provide information on extant/extinct ranges, native/introduced ranges, and seasonality.
-->

## Representative maps from each quadrant

@@

>  __Fig. S2.__ Breakdown of quadrants (Fig. S1A) by IUCN extinction risk categories.  IUCN and Aquamaps ranges were in better agreement (i.e., had a higher probability of being in the upper-right, "well-aligned" quadrant, Figure 2A/Figure S1A) for species at greater risk according to IUCN risk categories (p < 0.001, generalized linear model, family = binomial, Table S1).  This suggests at-risk species may receive more attention in terms of data collection or model refinement.

> __Table S1.__  Generalized linear model (family = binomial) describing how the agreement between IUCN and Aquamap range maps (i.e., probability of occurring in quadrant 1, Figure 2A/S1A) increases for a species as its IUCN risk status increases.  For this analysis, we excluded all species that had "data deficient" IUCN status (n = 255).  We analyzed the IUCN risk categories as a continuous variable (LC = 0.0, NT = 0.2, VU = 0.4, EN = 0.6, CR = 0.8, EX = 1.0).

_Model: glm(quadrant 1 membership ~ IUCN category, family = binomial) n = 2,059_

| Coefficients:  | Estimate | Std. Error | z value | p value      |
| :-----         |   -----: |     -----: |  -----: |       -----: |
| Intercept      | -1.32933 |   0.05804  | -22.902 | < 2e-16 ***  |
| IUCN category  |  1.18672 |   0.33469  |   3.546 | 0.000391 *** |

### Coral range vs. depth

@@

>  __Fig. S3.__ IUCN-defined range of _Oculina varicosa_, a typical photosynthetic coral, plotted against a 200 m bathymetry contour. According to IUCN, _O. varicosa_ can be found to depths of 152 m.  Maps of many other coral species and other reef-associated organisms followed similar patterns.

### AquaMaps threshold examination
